---
title: "Worksheet 2"
author: ""
date: "14.12.2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE)
```

## Environmental values and environmental education

Load the environmental values dataset (Pinder et al, 2020, https://doi.org/10.5061/dryad.wdbrv15kn).

```{r eval = FALSE}
env = read.csv("data/env_values.csv")
```

University students in Australia were asked various questions about their environmental values, desires for conservation careers, and educational background. We will examine two variables to address the hypothesis that students from families that value the environment rate more highly the quality of their education about environmental values.

The data has two variables:

* `env_education` contains responses to the question, "Thinking back to your schooling overall, how adequate was your education about environmental problems?", ranging from 0 (extremely inadequate) to 5 (extremely adequate).
* `value_of_env_protection` has responses for, "How important is protecting the environment in your family?", ranging from 0 (not important) to 7 (very important)... note that one student indicated -1 ;-)

1. Produce a histogram for each variable and a scatterplot for both variables. How do you interpret these plots? 
2. Which type of correlation is most appropriate for these two variables?
3. Use the function `cor.test` on `env_education` and `value_of_env_protection`. Is there a significant correlation? What direction is it? What is the estimate of the correlation?

**Hint**: Choose the type of correlation using `method` in `cor.test`. There are missing values in the data, you will need to deal with these. 


## Hydropsyche width-mass relationship

The following dataset collects head capsule widths and body masses from Hydropsyche (a genus of caddisflies) in the Danube in Austria. Note that this data file is tab delimited, not comma delimited, so you will need `read.table`.

```{r eval = FALSE}
hydrop = read.table("data/Hydropsyche.txt", header = TRUE)
```

1. As before, do some data exploration; plot the variables against each other, examine histograms, and possibly boxplots as well. How do you interpret these plots? Can you say anything about the relationship between these variables?
2. Use `lm` to fit a simple linear model (one variable, no transformations), using body mass as the response (y) variable, and width as the predictor. Is the regression significant? Report the statistics as shown in the lecture.

```{r eval = FALSE}
# use the formula syntax within lm to describe the model you want to fit
mod1 = lm(weight ~ width, data = hydrop)

# use the summary function to get information from your model
# how do you interpret the output?
summary(mod1)
```


3. Produce some diagnostic plots of your model (see slide 18: Regression in R: diagnostics, or use `plot(mod)`). Is this an adequate model? Does it meet the assumptions of the linear model?
4. Based on what you learned in 3, produce a new model, this time using a transformation of $x$ and/or $y$. Repeat 2 and 3 for the new model. Is the new model better? What does the new model imply about the relationship between mass and size? Is it testing the same hypothesis as the previous model?

**Bonus**: Produce a scatterplot of width (x-axis) and mass (y-axis), on the original scale with no transformations. Add the regression curve from question 3 to your plot.

```{r eval = FALSE}
# to get you started
ggplot(hydrop, aes(x = ?, y = ?)) + geom_point() + 
	geom_smooth(method = "lm", formula = ?)
```


